---
title: "Assignment 5: Data Visualization"
author: "Sarah Kear"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)
library(ggthemes)

#2 
processed_data = "Data/Processed_KEY"
PeterPaul.chem.nutrients <- read.csv(
  here(
    processed_data,
    "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE
  )

Niwot.Ridge.Litter <- read.csv(
  here(
    processed_data,"NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE
  )
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
A05_theme <- theme_base() + theme(
    plot.title = element_text( #Updating Title
      color= 'darkslategray',
      size = 14
    ),
    axis.text = element_text( #Updating axis text
      color = "black",
      size = 10
    ),
   axis.title.x = element_text( #Updating x-axis
     color = "black",
     size = 12
   ),
   axis.title.y = element_text( #Updating y-axis
     color = "black",
     size = 12
   ),
    legend.position = "top", #Putting legend to top
     legend.title = element_text( #Updating legend
      color='black',
      size = 12
    ),
   legend.text = element_text( #Updating legend text
     size = 12
     ),
    plot.background = element_blank() #takeaway plot edge
   )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
plot4 <- PeterPaul.chem.nutrients %>%
  ggplot(aes(
      x=po4,
      y=tp_ug,
      color=lakename)
    )  +
  geom_point(alpha=0.70) +
  xlim(0, 10) + # is this the correct way to hide the extreme values? should i keep it at 5, 10, 25?
  ylim(0, 75) +
  ylab("Total Phosphorus") +
  xlab("Total Phosephate") +
  ggtitle("Phosphorus by Phosphate at Paul Lake and Peter Lake") +
  scale_color_manual(values = c("darkorange2", "cornflowerblue"), 
                     name = "Lake Name") +
  A05_theme
print(plot4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
A05_theme1 <- theme_base() + theme(
    plot.title = element_text(
      color= 'darkslategray',
      size = 12
    ),
    axis.text = element_text(
      color = "black",
      size = 8
    ),
   axis.title.x = element_blank(), #leaving x-axis blank
   axis.title.y = element_text(
     color = "black",
     size = 10
   ),
     legend.title = element_text(
      color='black',
      size = 8
    ),
   legend.text = element_text(
     size = 8
     ),
   axis.text.x = element_text(
     angle = 90, 
     vjust = 0.5, 
     hjust=1),
   legend.position = "top",
   plot.background = element_blank()
   )

PeterPaul.chem.nutrients$month_f <- factor(
  PeterPaul.chem.nutrients$month, 
  levels=1:12,
  labels = month.abb
  )

temp <- PeterPaul.chem.nutrients %>%
  ggplot(aes(
      x=month_f,
      y=temperature_C,
      color=lakename)
    )  +
  geom_boxplot() +
  ylab(expression(paste("Temperature", degree, "C"))) +
  ggtitle("Temperature, Phophorus, & Nitrogen by Month") +
  scale_color_manual(values = c("darkorange2", "cornflowerblue"), 
                   name = "Lake Name") +
  A05_theme1 +
  theme(legend.position = "NONE") +
  scale_x_discrete(drop=FALSE)
temp

tp <- PeterPaul.chem.nutrients %>%
  ggplot(aes(
      x=month_f,
      y=tp_ug,
      color=lakename)
    )  +
  geom_boxplot() +
  ylab("Total Phosphorus") +
    xlab("Month") +
  scale_color_manual(values = c("darkorange2", "cornflowerblue"), 
                     name = "Lake Name") +
  A05_theme1 +
  theme(legend.position = "NONE",
        axis.title.x = element_text(
     color = "black",
     size = 10)) +
  scale_x_discrete(drop=FALSE)
tp

tn <- PeterPaul.chem.nutrients %>%
  ggplot(aes(
      x=month_f,
      y=tn_ug,
      color=lakename)
    )  +
  geom_boxplot() +
  ylab("Total Nitrogen") +
  scale_color_manual(values = c("darkorange2", "cornflowerblue"), 
                     name = "Lake Name") +
  A05_theme1 +
  scale_x_discrete(drop=FALSE)
tn

plot5 <- plot_grid(temp,
                   tp, 
                   tn, 
                   nrow = 1,
                   align = "h",
                   axis = "tblr",
                   rel_widths = c(1.5, 1.5, 1.5)
                   )
print(plot5)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temp is fairly the same between both lakes throughout the year. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
plot6 <- Niwot.Ridge.Litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(
      x=collectDate,
      y=dryMass,
      color=nlcdClass)
    )  +
  geom_point(alpha=0.65) +
  ylab("Dry Mass") +
  xlab("Collection Date") +
  ggtitle("Needle Dry Mass by Collection Date") +
  scale_color_manual(values = c("darkmagenta", "black", "darkolivegreen4"),
    name = "NLCD Class") +
  A05_theme +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(plot6)

#7
plot7 <- Niwot.Ridge.Litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(
      x=collectDate,
      y=dryMass,
      color=nlcdClass))  +
  geom_point() +
  ylab("Dry Mass") +
  xlab("Collection Date") +
  ggtitle("Needle Dry Mass by Collection Date") +
  facet_wrap(vars(nlcdClass)) + 
  A05_theme +
  scale_color_manual(values = c("darkmagenta", "black", "darkolivegreen4"),
    name = "NLCD Class") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8))
print(plot7)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: