---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sarah Kear"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(here)
here()

NTL_LTER_Physics <- read.csv(
  here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE
  )

NTL_LTER_Physics$sampledate <- as.Date(
  NTL_LTER_Physics$sampledate,
                                  format = "%m/%d/%y"
                                  )
head(NTL_LTER_Physics$sampledate, 5)

NTL_LTER_Physics$year4 <- year(
  as.Date(NTL_LTER_Physics$sampledate
          ))
head(NTL_LTER_Physics$year4, 5)

NTL_LTER_Physics$month <- month(
  as.Date(NTL_LTER_Physics$sampledate
          ))
head(NTL_LTER_Physics$month, 5)

#2
A07theme <- theme_classic(base_size = 12) +
  theme(
    axis.text = element_text(
      color = "black",
      size = 10
      ),
   axis.title.x = element_text( #Updating x-axis
     color = "black",
     size = 12
   ),
   axis.title.y = element_text( #Updating y-axis
     color = "black",
     size = 12
   ),
    legend.position = "top", #Putting legend to top
     legend.title = element_text( #Updating legend title
      color='black',
      size = 12
    ),
   legend.text = element_text( #Updating legend text
     size = 10
     ),
    plot.background = element_blank() #removing plot edge/background
   )
theme_set(A07theme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no difference in mean lake temperature during July across all lakes' depth.
Ha: There is difference in mean lake temperature during July across all lakes' depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTER_Processed <- NTL_LTER_Physics %>%
  filter(NTL_LTER_Physics$month == 7) %>%
  select(lakename:daynum,
         depth:temperature_C
         ) %>%
  na.omit()
head(NTL_LTER_Processed, 5)

#5
plot1 <- NTL_LTER_Processed %>%
  ggplot(aes(
    x=depth,
    y=temperature_C
  )) +
  geom_point(alpha = 1/5, color = "azure4", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, col="slateblue2") +
  ylim(0, 35) +
  ylab(expression(paste("Temperature", degree, "C"))) +
  xlab("Depth (meters)") +
  ggtitle(expression(paste("Temperature", degree, "C by Depth (m)"))) +
  A07theme
print(plot1)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Overall, there seems to be a negative correlation between temperature and depth. Temperature decreases as the lake depth increases. However, there is a a fairly wide range in terms of the temperature distribution by depth. the points are not necessarily close to smooth line, illustrating that that relationship between temperature and depth is not be that linear.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
temp_lr1 <- lm(data = NTL_LTER_Processed,
               temperature_C ~ depth)
summary(temp_lr1)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: *Temperature = 21.96 - 1.95[depth]*. A meter increase in depth is associated with a 1.95 decrease of temperature, holding all else constant. Depth's regression coefficient has a p-value that is less than 0.05, meaning that depth has a statistically significant relationship with temperature. The adjusted r-squared is 0.7387, meaning that around 73.9% of the changes in temperature is due to changes in depth. The degree of freedom the finding is based off of is 9,726. We are able to reject the null hypothesis.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
Temp_AIC <- lm(data = NTL_LTER_Processed,
               temperature_C ~ year4 + daynum + depth)

step(Temp_AIC)

#10
summary(Temp_AIC)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests are the three variables included. Removing year, daynum, or depth would increase the AIC. The oberserved variance only explain 74% of change in temperature. This is a slight imporvement in adusted r-squared, but i don't think it's that much improvement as it now takes three explanatory variables in explain 74% of temperature change, when it only takes one explanatory variables (depth) that explains 73.9% of temperature change.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Do i need to create a sep. dataset that groups by lake name, year, and daynum and creates the average temp. for that day per year???
lake.july.ANOVA <- aov(data = NTL_LTER_Processed,
                       temperature_C ~ lakename)
summary(lake.july.ANOVA)

lake.july.ANOVA2 <-lm(data = NTL_LTER_Processed,
                       temperature_C ~ lakename)
summary(lake.july.ANOVA2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is significant difference in the mean temperature among the different lakes. We confirm this as the p-value is less than 0.05. We are able to reject the null hypothesis that all lake means are equal.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
plot2 <- NTL_LTER_Processed %>%
  ggplot(aes(
    x=depth,
    y=temperature_C,
    color=lakename
  )) +
  geom_point(alpha = .5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1, aes(group = lakename)) +
  ylim(0, 35) +
  ylab(expression(paste("Temperature", degree, "C"))) +
  xlab("Depth (meters)") +
  ggtitle(expression(paste("Temperature", degree, "C by Depth (m)"))) +
  guides(color=guide_legend(title="Lake", nrow=3, byrow=TRUE)) +
  A07theme
print(plot2)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lake.july.ANOVA)
lake.july.groups <- HSD.test(lake.july.ANOVA, "lakename", group = TRUE)
lake.july.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Using the HSD.test(), Paul Lake and Ward Lake have the same mean as Peter Lake. There is not a lake that has a mean temperature that is statistically distinct from any of the other lakes. All lakes are groups with other lakes with the same mean.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use the two-sample t-test to test for equal variances.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_LTER_Crampton_Ward <- NTL_LTER_Processed %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))
head(NTL_LTER_Crampton_Ward, 5)

temp.twosample <- t.test(NTL_LTER_Crampton_Ward$temperature_C ~ NTL_LTER_Crampton_Ward$lakename)
temp.twosample
```

>Answer: We cannot reject the null hypothesis, as the p-value is greater than 0.05. The mean from both lakes are statistically the same. The mean temperature amounts for each lake are the same as the lake's respetive outputs from Question 16.
